Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-16 implementation/implementation.tex
Grammar fix

According to our preliminary studies, a ABC SMC framework implemented using \verb|pyABC| in \verb|Python| was adopted in this project. \verb|pyABC| is a popular SMC software packages\cite{ref:pyabc} which has been used form many ABC SMC inference applications\cite{population}. \verb|pyABC| provides an open-source framework for likelihood-free inference, which is a \verb|NumPy| implementation of ABC SMC algorithm and a useful tool-box for our own inference and analysis tasks. Besides, it supports multi-core execution implemented using \verb|multiprocessing| built-in package, and it made our further studies in the scaling-up performance possible.

@@

The inference framework code for the project was developed and tested in local environment (macOS 10.15.6, see Table \ref{table:local_macine}) at first using a small input, then the functional version was deployed on compute node of ARCHER and Cirrus for parameter inference experiments and model selection tasks. The scaling-up experiment was performed on Cirrus. The profiling and its analysis were performed on local machine. Some necessary development and debug were performed on the remote machines to ensure that experiments run correctly on compute node.

The ODE model-related code and utilities were firstly developed to enable reusable functions, variables and data calls. It included the code for the ODE equations, ODE solver, data structures and data format transform functions, etc. By doing this in separate code files, we could change implementation options and activate ABC SMC easily in another `trigger' file, without creating definition and duplicated code for the models, observed data and parameter priors.

Some missing features were found when using \verb|pyABC|, therefore some modifications were made to the source code of \verb|pyABC|.

For example, the laboratorial measured data were not considered `complete' or `tidy' for modeling and analysis purpose: the available time points for cells and cytokines were different, e.g. cell number was not measured at 0.25 hpl and cytokines' expression was not measured at time point 120 hpl. These missing values were treated as `ignored' when calculating the distance between observed data and simulated data, and represented by \verb|NaN|s. However, in some distance calculations \verb|NaN| was not acceptable; to cope with this, built-in distance function was modified. Other modified code include some visualisation and result representation enhancements

Analysis of the program output was also wrapped up into functions which can be easily called after an ABC SMC run is finished, which includes multiple visualisations and summary statistic of the results. For different tasks there are also separated analysis code file. Other analysis tools used in the project include built-in database visualisation tool, R code and Microsoft Excel.

@@

After several development iterations, code files for ABC SMC implementation now mainly had two types: code files for infer-back experiments (using synthetic data), and code files for parameter inference (using experimental data). Model selection can be easily done by adding more model to the implementation code file.

@@

As the key focus of this project was on the parameter estimations of dynamical system, the ability of inference on the proposed models was firstly examined before actual inference using the experimental observed data.

In the first part, synthetic data is used with known true parameter values. The algorithm implementation is evaluated by the efficiency and goodness of resultant model under different implementation options and ABC SMC hyperparameters.

Then according to findings in the first part, several options with good performance are tried in the inference using experimental observed data (Figure \ref{fig:obs_data}). To obtain a more accurate and general results, these experiments are repeated for 3 or more times.

@@

The ABC SMC inference can have largely different performance when using different hyperparameters and implementation options. To firstly test the ability of inference and observe the results of different options, synthetic data with known true parameter values are used as target data. by doing this, we could directly compare the inferred parameters to the true value, and compare the simulated trajectories to the true trajectories and then observe the result to see whether the inference is successful and how well the model fit the data. Also, by setting the same final threshold value, the efficiency under different options can be compared and help us to choose proper options when applying ABC SMC on the observed data. The true value of parameters that were used to generate synthetic data is listed in Table \ref{table:known_values}.

The true values were obtained by fitting model 1 onto the experimental data (Figure \ref{fig:infer_back_data}) using least-square optimisation method. Available least-square fit tools in \verb|SciPy| were used. The fit algorithm was Trust Region Reflective\footnote{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.optimize.least_squares.html}}. Target data was then generated using these parameter values via ODE integrate solver\footnote{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.integrate.odeint.html}}.

@@

The following topics were studied using the synthetic data and ABC SMC inference framework.

@@

Perturbation kernels work in the sampling process. In each generation $t$, samples are firstly taken from the previous population $\{\theta_{t-1}\}$ with weights $\{w_{t-1}\}$, then perturbed using the perturbation kernel $K(\theta|\theta^*)$. We keep sampling until $N$ particles are accepted, where $N$ is the pre-set population size. After that the new weights are calculated and normalised. Figure \ref{fig:smc} illustrates the process.

% \begin{figure}[t!]
%     \begin{center}
%         \resizebox{1.0\hsize}{!}{\includegraphics{fig/smc.png}}
%     \end{center}

%     \caption[ABC SMC sampling process]%
%     {ABC SMC sampling process. $\theta^*$ denotes a sample drawn from previous population, which is used to generate simulated data $D^*$. $d$ is a distance measurement, $d(D^*,D)$ represents the discrepancy between observed data and simulated data. for each population $t$, $\epsilon_t$ is a threshold criterion to determine whether to accept that drawn sample}
%     \label{fig:smc}

The perturbation kernel is called in the sampling of every particles and determines how the new perturbed particle is chosen, thus it is influential to both computation complexity and the resultant posterior distribution. Generally, a local perturbation kernel may face the risk of being stuck in local modes (e,g., local optimal), but it may need less computational operations, or could generate a population with a higher acceptance rate if the successive epsilon values are close; A kernel with wide variance, or spreading out in a large space could help in resolving the local optimal problem by a more thoroughly exploration of the parameter space, however it can be more computation-intensive and result in a lower acceptance rate. A desired optimal kernel should balance the trade-offs; their property and criteria is discussed in \cite{ref:kernel}.

There are several common choice of perturbation kernels. Among them multivariate normal kernel and local M-nearest neighbour model is preferred to be applied on our models. A covariance matrix $\Sigma_t$ of accepted particles is calculated form previous generation and used in multivariate normal kernel: $K(\theta|\theta_{t-1})\sim\mathcal{N}(\theta_{t-1}, \Sigma_t)$. It is illustrated to be more efficient than uniform kernel and component-wise normal kernel in relecting the true posterior structure \cite{ref:kernel}. It has been proved to perform well in several dynamic system models \cite{ref:abcsysbio, ref:compare, ref:disease} for the parameter estimation and model selection tasks. The $scaling\in(0,1]$ parameter in \verb|pyABC| will be multiplied to the covariance to produce a `narrower distributed' perturbation result.

Local M-NN kernel provided by \verb|pyABC| is also tried to provide a comparison. A local kernel density estimation (KDE) fit is used with M nearest neighbours considered.

The kernel experiments is designed to explore the efficiency of SMC on our dynamical systems. Given the same fixed threshold schedule, kernels that need less total samples and have higher acceptance rate would be our preference. The experiments compared multivariate normal kernel with local MNN kernels with different parameters, using synthetic data to infer back the parameters. The acceptance rate among each time point and total required samples are compared after the experiment to give suggestions on the kernel selection in the real data inference. As the threshold schedule is fixed, the final population should have similar discrepancy to the target data thus here the goodness of fit i.e. recover of target data trends/features and errors of the inferred parameters compared to true values is not discussed.

@@

To compare kernels, a fixed schedule schedule is used with minimal epsilon i.e. $\epsilon_t$ set to 10.0. From the total required samples graph Figure \ref{fig:kernel1}, the efficiency can be compared. Among the tested kernels, the local M-NN with M=50 has the best performance: it requires the least number of samples and
has higher acceptance rates among almost all generations. Local M-NN with M=750 is the slowest kernel.

For local M-NN kernels, generally a greater M will lead to lower acceptance rate and
more required particles in each generation. As shown in Figure \ref{fig:acceptance1}, local M-NN with M=750 is the lowest curve. Consequently, if greater M is test, then it will have even lower acceptance rates; the maximum M=2000 (whole population is considered) will have the lowest acceptance rates.

@@

Multivariate normal kernel has a performance between local M-NN M=250 and M=100. It proves that rather than a trivial normal kernel, multivariate normal kernels is more efficient facing the concentrations of joint distributions among multiple parameters \cite{ref:kernel}. `Scaling' option can narrower the distribution calculated from the last population, thus making the kernel more `local' by sampling under a smaller variance. Similar to small M in local M-NN, small `scaling' is also more efficient in our experiment of the model 1. Besides the fixed threshold schedule, a median threshold schedule with the same final threshold $\epsilon_{20}=10.0$ is also tested and similar results are observed (Figure \ref{fig:kernel2} and \ref{fig:acceptance2}, Appendix B2).

A more `local' kernel was more likely to give better performance by efficiently sampling around concentrations in parameters' distribution and approximate the posterior distribution. This holds in most cases under a given target final threshold value $\epsilon_t$, however may not produce the proper result want: a more local kernel, e.g. local M-NN with small M and multivariate normal with `scaling' is more likely to be stuck in local optimality, as the kernel is less likely to sample particles that are far from local concentrations, although these local optimal are still accepted under given threshold. In other words, using local kernels the epsilon may quickly converge to to local optimal; if a even smaller epsilon is desired, the local kernel can hardly generate enough particles to find another matched local modes. Also, the shapes of posterior distribution will affect the performance of local kernels \cite{ref:kernel}. One obvious example is presented in Figure \ref{fig:kernel1}, where local M-NN with M=50 suffered from a local optimal: the last generation takes much more samples to meet the required threshold.

\subsubsection{Conclusion} Local kernels can be fast bur unstable in some case; for a general parameter inference of our models, multivariate kernels are preferred, as a good fit is our prior target; some local kernels are also worth trying after the multivariate normal kernel, e.g. local M-NN with M$\leq 250$ (for population size of 2000, i.e. 12.5\% of the population) and multivariate normal with scaling. Also, multivariate M-NN is worth trying, although it was not built in \verb|pyABC| and requires additional implementations.

@@

where $i$ is the index of data points and $\Delta x_i$ is the discrepancy between  observed data and simulated data at data point $i$, i.e. $\Delta x_i = x_{i, simulated}-x_{i, observed}$. In this case data points are 12 time points of four variables i.e. 48 data points in total.

@@

where a weight $w_i$ is assigned to every data point. If data point $i$ have a higher weight, then the data value is regeared to be more informative, i.e. giving more help in inferring the true posteriors; weights can be either pre-set according to prior knowledge of the problem, or using adaptive method to be dynamically calculated according to , known as adaptive distance.

@@

Factor is helpful as an option for efficiency. When some data points are equally informative, the manually pre-defined can make the distance more focused on certain data point, or behave like a normalisation that can balance the scales of different summarised statistics (here is the 48 mean values). There are two appliance case studied: (1) factors used as an normalisation option and (2) factors used to give more focus on some features of the observed data. For (2), as we noticed that the main features (e.g. rapid increase, peaks, fluctuations) are mostly observed in the first half of the time points i.e. 0 - 30 hpl, factors are tried in the later parameter estimation of real data where a more accurate fit of the curve is desired.

Adaptive population \cite{population} can adapt the population size of each generation according to the mean correlation of variation error of the previous population. In our early test the maximal population size that is allowed is set to 5,000 and the adaptive strategy always select the upper bound, as a results of wide parameter space or the wide posterior approximation shape.

@@

Two types of factor and adaptive distance function were tested by running the same number of generations (20 generation, 2000 particles per generation). The first type of factor is 25:75  factor, where first half of the trajectory data was assigned higher factor (75), and second half of the data was assigned lower factor (25); this could make the first half if the data more `important' in the inference. The second factor was range factor, which tried to normalise the trajectory of each dependent variable according to its data range.

From Figure \ref{fig:factor}, different types of factor gave different performance. 25:75 factor required much more samples in the first generation (orange bar in the left of Figure \ref{fig:factor}) and also did not improve much in later generations, thus it is the slowest one. The first half of observed data contained more features and information, assigning it with higher factor would make the algorithm think that it is more important to fit the first half and thus focus on the hard part, and consequently more particles were needed in the inference process. The acceptance rates of 25:75 factor among different generations were also generally slightly lower than the standard. By contrast, range factor required less total samples and gave generally higher acceptance rates than standard, which made it considerably faster. Adaptive distance gave the best efficiency, as it significantly improved the acceptance rates and resulted in even less required samples (right, Figure \ref{fig:factor}).

@@

However, the result of factors adaptive distance could not be directly compared to standard run, because it changed the distance measurement by adapting weights or manually setting factors. Factors were normalise to make it more close the standard distance measurement. As a result, the same final threshold value did not guarantee that adaptive distance can reach the same level of fit as others. After 20 populations, adaptive distance had a poor fit (Figure \ref{fig:factor_sim}), while the standard run and two types of factors had similar tight fits of the data, which all converge to the true posterior. Additional in our implementations, variance factor (which tries to normalise the data according to the dependent variables' variance) was also tries and gave very similar performance as range factor.

To conclude, because the factors and weights are directly multiplied to $\Delta x$ and thus the metrics for distance are changed, the direct compare of the efficiency under the same threshold schedule was unfeasible; we could only compare the results after the same number of generations. Factors gave nearly the same inferred results but required different number of samples, hence cares must betaken when choosing factor schemes. Adaptive distance did not result in a accurate result as others. As a result, our further implementations used standard distance function, and factors were tried for possible improvements.

@@

Regarding the existing dynamic systems model, SMC was applied with two data size options, three prior distribution (uniform distribution, wider uniform distribution and log-uniform distribution). The population options i.e. population size and number of populations were also tried with different values. These experiments intended to give suggestions on the later SMC implementation on experimental data.

@@

The results from data size and prior distribution range is shown in Figure \ref{fig:size}. The standard implementation is fed with 120 data points (30 data points for each of the four variables), the `less data' is fed with 48 data points, which is the case of the real experimental measurement. Although much less (60\%) data was used, the total required sampling numbers was only 4.47\% less. This indicated that the data size does not affect the sampling process much and the inferred model are all considered acceptable when compared to the synthetic data, as they finally reached a similar epsilon threshold ((b), Figure \ref{fig:size}), but cares should still be taken when using less data point, or using other more summative statistics (e.g. mean, standard deviation, peak values, starting and ending values etc.) as the target data: as shown in \cite{ref:disease}, less data can lead to a model with more variance, under-fitting or missing some local features such as local peaks.

The prior range and distribution can largely affect the inference progress (Figure \ref{fig:size}. (a)). Samples were drawn from a high dimensional parameter space in each sampling process, a wider parameter prior interval in only one dimension can cause the parameter space to be much bigger, and consequently the exploration and sampling of parameter vectors would take much more efforts. Figure \ref{fig:size}. (b) also indicates that the convergency is slower for wider prior ranges. As a result, a proper prior interval based on our experience and belief with seeing the data should be set as narrow as possible, as inference are largely affected by the prior interval and its length; improper prior could even make the algorithm miss the true posterior \cite{ref:abcsysbio}.

How to specify prior can be tricky from problem to problem. For our task in the parameter inference of regeneration model, biologically motivated priors were used, which were general reflection of biological constrains. Also the units of the parameters help us in identify the possible scale of the parameters.

@@

This subsection aims to preliminarily explore options in ABC SMC implementations. From the above results, we adopted multivariate normal kernel as the kernel for our real inference on observed data; Local M-NN kernels were also planned to be tested. In the real inference, 25:75 factor were also tried to see if it could help in recovering more features and resulting a better fit. because the improvement by using less data was inapparent, we did not try to change the target summary data size. Prior intervals were explored explicitly and gradually shrunk for our models.

Some other hyperparameter, e.g. number of generations, population size (i.e. number of particles per generation) were not addressed in this subsection. In the real inference, population size was set to be greater than 2000 to avoid insufficient exploration of the previous population; in some case than high variance or uncertainties in the results were observed, population size as large as 20,000 were used. If the epsilon did not show a convergency trend, the run could be extended to more generations until a stable result was observed.

@@

Using the suggested options in the previous section, the target data (Figure \ref{fig:obs_data}) is prepared as input for the SMC inference framework. Regarding the result, the population size and number of generations are adjusted several times to obtain an general informative posterior which should be stable across repeated runs and avoid local optimal as much as possible.

Parameter estimation of model 1, 2 and 3 was tried with different prior distribution setting: distribution range [0, 25] and [0, 75], amd distribution type uniform and log-uniform. All these ABC SMC runs uses 2,000 particles per population and 30 generations.

Log-uniform-shaped prior seems to give a better fit of the model, and wider prior range [0, 75] does not give a more accurate result. Then distribution range [0, 25] was tried, in case some parameter have a true posterior laying between [25, 50].

The results of prior distribution log-uniform [0, 50] is shown in Figure \ref{fig:result123}. For model 3, the acceptance rates and epsilon path are presented in Figure \ref{fig:result123_2}. For model 3, the estimated parameters' posterior is shown in Figure \ref{fig:para1}. Figure \ref{fig:result123} plots the estimated curve of models, using the mean value of each parameter's approximated posterior; also 1000 particles (each particle is a parameter set) were sampled and used to generate simulated trajectories of the four variable, and the mean (blue) and 25th to 75th percentile range (grey) are calculated from the 1000 simulated data.

Before applying model comparison methods, some features of the three model can still be compared. From simulated curves, It can be seen that all the inter-quartile ranges are tight around the mean value curve, which indicates the approximated posterior are concentrated to some degree; the peak-shaped posterior distributions (Figure \ref{fig:para1}) agrees with this. All epsilon values converges to a steady level, and model 3 has a lower final epsilon value and consequently the simulated trajectory is more close to the observed data (model 3, Figure \ref{fig:result123}).

The acceptance rates is fluctuating and have a gradually increase trend for all the three models, as the acceptance rate becomes higher when the true posteriors are gradually approximated. Compared to model 1, model 2 gives a better fit of the decreasing trend at the second half of the observed data, by using a exponentially decaying self-increase rate $\lambda_N$ instead of a constant. The resultant simulated data from model 2 and 3 are in the similar trends and the most obvious difference is that for model 3, the simulated data of $\Phi$ and $\alpha$ are more `flat'. We expect model 3 to be the best model as it reaches a threshold value and intuitively gives a better fit, although the fit could not be considered generally a good representation of the biological process that we want to model, as the simulated data are significantly biased from the observed data for tnf-$\alpha$. A sharp peak at 4 hpl is observed from the measurement (Figure \ref{fig:obs_data}) but no similar features are represented by any of the three models.

@@

[model comparison among model 1, 2, 3]

>>>>

@@

According to our preliminary studies, an ABC SMC framework implemented using \verb|pyABC| in \verb|Python| was adopted in this project. \verb|pyABC| is a popular SMC software packages\cite{ref:pyabc} which has been used form many ABC SMC inference applications\cite{population}. \verb|pyABC| provides an open-source framework for likelihood-free inference, which is a \verb|NumPy| implementation of ABC SMC algorithm and a useful tool-box for our inference and analysis tasks. Besides, it supports multi-core execution implemented using \verb|multiprocessing| built-in package, and it made our further studies in the scaling-up performance possible.

The inference framework code for the project was developed and tested in a local environment (macOS 10.15.6, see Table \ref{table:local_macine}) at first using a small input, then the functional version was deployed on compute node of ARCHER and Cirrus for parameter inference experiments and model selection tasks. The scaling-up experiment was performed on Cirrus. The profiling and its analysis were performed on a local machine. Some necessary development and debug were performed on the remote machines to ensure that experiments run correctly on the compute node.

The ODE model-related code and utilities were first developed to enable reusable functions, variables and data calls. It included the code for the ODE equations, ODE solver, data structures and data format transform functions, etc. By doing this in separate code files, we could change implementation options and activate ABC SMC easily in another `trigger' file, without creating definition and duplicated code for the models, observed data and parameter priors.

Some missing features were found when using \verb|pyABC|; therefore some modifications were made to the source code of \verb|pyABC|.

For example, the laboratory-measured data were not considered `complete' or `tidy' for modelling and analysis purpose: the available time points for cells and cytokines were different, e.g. cell number was not measured at 0.25 hpl and cytokines' expression was not measured at time point 120 hpl. These missing values were treated as `ignored' when calculating the distance between observed data and simulated data, and represented by \verb|NaN|s. However, in some distance calculations, \verb|NaN| was not acceptable. To cope with this, the built-in distance function was modified. Other modified code included some visualisation and result representation enhancements.

Analysis of the program output was also wrapped up into functions which can be easily called after an ABC SMC run is finished, which includes multiple visualisations and summary statistic of the results. For different tasks, there are also separated analysis code file. Other analysis tools used in the project include built-in database visualisation tool, R code and Microsoft Excel.

@@

After several development iterations, code files for ABC SMC implementation now mainly had two types: code files for infer-back experiments (using synthetic data), and code files for parameter inference (using experimental data). Model selection can be easily performed by adding more model to the implementation code file.

@@

As the key focus of this project was on the parameter estimations of a dynamic system, the ability of inference on the proposed models was firstly examined before actual inference using the experimental data.

In the first part, synthetic data (Table \ref{table:known_values}) was used with known true parameter values. The algorithm implementation was evaluated by the efficiency and goodness of the resultant model under different implementation options and ABC SMC hyperparameters.

Then according to findings in the first part, several options with considerable performance were tried in the actual inference with observed data (Figure \ref{fig:obs_data}). To obtain more accurate and general results, these experiments were repeated for three or more times.

@@

    The ABC SMC inference can have widely different performance when using different hyperparameters and implementation options. To firstly test the ability of inference and observe the results of different options, synthetic trajectories with known true parameter values were used as target data. By doing this, we could directly compare the inferred parameters to the true values and compare the simulated trajectories to the true trajectories, then observe the result to see whether the inference is successful and how well the model fit the data. Also, by setting the same final threshold value, the efficiency under different options can be compared and help us to choose proper options when applying ABC SMC.

    The true values (Table \ref{table:known_values}) were obtained by fitting model 1 onto the experimental data (Figure \ref{fig:infer_back_data}) using least-square optimisation method. Available least-square fit tools in \verb|SciPy| were used. The fitting algorithm was Trust Region Reflective\footnote{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.optimize.least_squares.html}}. Target data was then generated using these parameter values via ODE integrate solver\footnote{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.integrate.odeint.html}}.

@@

The following topics were studied using the synthetic data and the developed ABC SMC inference framework.

@@

Perturbation kernels work in the sampling process. In each generation $t$, samples are firstly taken from the previous population $\{\theta_{t-1}\}$ with weights $\{w_{t-1}\}$, then perturbed using the perturbation kernel $K(\theta|\theta^*)$. We keep sampling until $N$ particles are accepted, where $N$ is the pre-set population size. After that, the new weights are calculated and normalised. Figure \ref{fig:smc} illustrates the process.

The perturbation kernel is called in the sampling of every particle and determines how the new perturbed particle is chosen; thus it is influential to both computation complexity and the resulting posterior distribution. Generally, a local perturbation kernel may face the risk of being stuck in local modes (e,g., local optimal), but it may need less computational operations or could generate a population with a higher acceptance rate if the successive epsilon values are close. A kernel with a wide variance, or spreading out in a large space could help in resolving the local optima problem by a more thorough exploration of the parameter space. However, it can be more computation-intensive and result in a lower acceptance rate. A desired optimal kernel should balance the trade-offs; their properties and criteria were discussed in \cite{ref:kernel}.

There are several common choices of perturbation kernels. Among them, the multivariate normal kernel and local M-nearest neighbour model is preferred to be applied to our models. A covariance matrix $\Sigma_t$ of accepted particles is calculated from the previous generation and used in multivariate normal kernel: $K(\theta|\theta_{t-1})\sim\mathcal{N}(\theta_{t-1}, \Sigma_t)$. It is illustrated to be more efficient than uniform kernel and component-wise normal kernel in reflecting the true posterior structure \cite{ref:kernel}. It has been proved to perform well in several dynamic system models \cite{ref:abcsysbio, ref:compare, ref:disease} for the parameter estimation and model selection tasks. The $scaling\in(0,1]$ parameter in \verb|pyABC| will be multiplied to the covariance to produce a `thinner-distributed' perturbation result.

Local M-NN kernel provided by \verb|pyABC| was also tried to provide a comparison. A local kernel density estimation (KDE) fit is used with M nearest neighbours considered.

The kernel experiment is designed to explore the efficiency of SMC on our dynamical systems. Given the same fixed threshold schedule, kernels that need less total samples and have higher acceptance rates would be our preference. The experiments compared multivariate normal kernel with local MNN kernels with different parameters, using synthetic data to infer back the parameters. The acceptance rate among each time point and total required samples are compared after the experiment to give suggestions on the kernel selection in the real data inference. As the threshold schedule was fixed, the final population should have similar discrepancy to the target data thus here the goodness of fit, i.e. recover of target data trends/features and errors of the inferred parameters compared to true values were not discussed.

@@

To compare kernels, a fixed schedule was used with the final epsilon $\epsilon_t=10.0$. From the total required samples plot (Figure \ref{fig:kernel1}) the efficiency can be compared. Among the tested kernels, the local M-NN with M=50 has the best performance: it requires the least number of samples and
has higher acceptance rates among almost all generations. Local M-NN with M=750 is the slowest kernel.

For local M-NN kernels, generally, a higher M leads to lower acceptance rates and more required particles in each generation. As shown in Figure \ref{fig:acceptance1}, local M-NN with M=750 is the lowest curve. Consequently, if greater M was tested, then it would have even lower acceptance rates; the maximum M=2000 (the whole population is considered) would have the lowest acceptance rates.

@@

Multivariate normal kernel has a performance between local M-NN M=250 and M=100. It proves that rather than a trivial normal kernel, multivariate normal kernels are more efficient facing the concentrations of joint distributions among multiple parameters \cite{ref:kernel}. `Scaling' option can narrow the distribution calculated from the last population, thus making the kernel more `local' by sampling under a smaller variance. Similar to small M in local M-NN, small `scaling' is also more efficient in our experiment of model 1. Besides the fixed threshold schedule, a median threshold schedule with the same final threshold $\epsilon_{20}=10.0$ was also tested and similar results are observed (Figure \ref{fig:kernel2} and \ref{fig:acceptance2}, Appendix B2).

A more `local' kernel is more likely to give better performance by efficiently sampling around concentrations in parameters' distribution and approximate the posterior distribution. This holds in most cases under a given target final threshold value $\epsilon_t$. However, it may not produce the proper result want: a more local kernel, e.g. local M-NN with small M and multivariate normal with `scaling' is more likely to be stuck in local optimality, as the kernel is less likely to sample particles that are far from local concentrations, although these local optimal are still accepted before even smaller threshold values is demanded (see Section 5.2 for more discussions). One obvious example is presented in Figure \ref{fig:kernel1}, where local M-NN with M=50 suffered from a local optimal: the last generation takes much more samples to meet the required threshold.

% In other words, by using local kernels the epsilon may quickly converge to a local optimum; if an even smaller epsilon is desired, the local kernel can hardly generate enough particles to find another matched local modes.

% Also, the shapes of posterior distribution affect the performance of local kernels \cite{ref:kernel}.

To conclude, Local kernels can be fast but unstable in some case. For a general parameter inference of our models, multivariate kernels were preferred, as a good fit is our prior target. Some local kernels are also worth trying after the multivariate normal kernel, e.g. local M-NN with M$\leq 250$ (for a population size of 2000, i.e. 12.5\% of the population) and multivariate normal with scaling, but extra cares must be taken: although they are fast, moving in and moving out of local optimum make them more unstable in efficiency.

% Also, multivariate M-NN is worth trying, although it was not built in \verb|pyABC| and requires additional implementations.

@@

where $i$ is the index of data points and $\Delta x_i$ is the discrepancy between  observed data and simulated data at data point $i$, i.e. $\Delta x_i = x_{i, simulated}-x_{i, observed}$. In this case, data points are 12 time points of four variables, i.e. 48 data points in total.

@@

where a weight $w_i$ is assigned to every data point. If data point $i$ has a higher weight, then the data value is regeared to be more informative, i.e. giving more help in inferring the true posteriors. Weights can be either pre-set according to prior knowledge of the problem or updated using an adaptive method. This method is known as adaptive distance.

@@

Factor is helpful as an option for efficiency. When some data points are equally informative, the manually pre-defined can make the distance more focused on certain data points, or behave like a normalisation that can balance the scales of different summarised statistics (here are the 48 mean values). There are two appliance case studied: (1) factors used as a normalisation option and (2) factors used to give more focus on some features of the observed data. For (2), as we noticed that the main features (e.g. rapid increase, peaks, fluctuations) are mostly observed in the first half of the time points, i.e. 0 - 30 hpl, factors were tried in later experiments. Such factor scheme assigns data in the first half of the time point with higher factors (75) and data in the second half with lower factors (25).

Adaptive population \cite{population} can adapt the population size of each generation according to the mean correlation of variation error of the previous population. However, in our early tests,  the maximal allowed population size is set to 10,000, and the adaptive strategy always selected the upper bound (10,000) for new populations, as a result of wide parameter space or the wide posterior approximation shape. Additionally, some internal bugs were observed when applying it. Hence, the adaptive population was not tried in this project; however, it is worth trying in future works as an efficiency improvement.

@@

Two types of factor and adaptive distance function were tested by running the same number of generations (20 generations, 2000 particles per generation). The first type of factor is 25:75  factor, where the first half of the trajectory data were assigned higher factor (75), and the second half of the data was assigned lower factor (25); this could make the first half of the data more `important' in the inference. The second factor was `range' factor, which tried to normalise the trajectory of each dependent variable according to its data range.

From Figure \ref{fig:factor}, different types of factor gave different performance. 25:75 factor required much more samples in the first generation (orange bar in the left of Figure \ref{fig:factor}) and also did not improve much in later generations; thus, it is the slowest one. The first half of observed data contained more features and information, assigning it with higher factor would make the algorithm think that it is more important to fit the first half and thus focus on the hard part, and consequently more particles were needed in the inference process. The acceptance rates of 25:75 factor among different generations were also generally slightly lower than the standard. By contrast, range factor required less total samples and gave generally higher acceptance rates than standard, which made it considerably faster. The adaptive distance gave the best efficiency, as it significantly improved the acceptance rates and resulted in even less required samples (right, Figure \ref{fig:factor}).

@@

However, the result of factors adaptive distance could not be directly compared to standard run, because it changed the distance measurement by adapting weights or manually setting factors. Factors were normalised to make it closer to the standard distance measurement. As a result, the same final threshold value did not guarantee that adaptive distance can reach the same level of fit as others. After 20 populations, adaptive distance had a poor fit (Figure \ref{fig:factor_sim}), while the standard run and two types of factors had similar tight fits of the data, which all converge to the true posterior. Additional in our implementations, variance factor (which tries to normalise the data according to the dependent variables' variance) was also tried and gave a very similar performance as the range factor.

To conclude, because the factors and weights are directly multiplied to $\Delta x$, and thus the metrics for distance are changed, the direct comparison of the efficiency under the same threshold schedule was unfeasible; we could only compare the results after the same number of generations. Factors gave nearly the same inferred results but required a different number of samples, hence cares must be taken when choosing factor schemes. Adaptive distance did not provide an accurate result as others. As a result, our further implementations used standard distance function, and factors were also tried for possible improvements.

@@

Regarding the existing dynamic systems model, SMC was applied with two data size options and two ranges of the prior interval. The population options, i.e. population size and the number of populations were also tried with different values. These experiments intended to give suggestions on the later implementation on experimental data.

@@

The results from data size and prior distribution range are shown in Figure \ref{fig:size}. The standard implementation is fed with 120 data points (30 data points for each of the four variables), the `less data' is fed with 48 data points, which is the case of the real experimental measurement. Although much less (60\%) data was used, the total required sampling numbers was only 4.47\% less. This indicated that the data size does not affect the sampling process much and the inferred model are all considered acceptable when compared to the synthetic data, as they finally reached a similar epsilon threshold (Figure \ref{fig:size}), but cares should still be taken when using less data points, or using other more summative statistics (e.g. mean, standard deviation, peak values, starting and ending values etc.): as shown in \cite{ref:disease}, fewer data can lead to a model with more variance, under-fitting or missing some local features such as local peaks.

The prior range and distribution can largely affect the inference progress (Figure \ref{fig:size}. (a)). Samples were drawn from a high dimensional parameter space in each sampling process. A wider parameter prior interval in only one dimension can cause the parameter space to be much bigger. Consequently, the exploration and sampling of parameter vectors would take much more efforts. Right plot in Figure \ref{fig:size} also indicates that the convergency is slower for wider prior ranges. As a result, a proper prior interval based on our experience and belief with seeing the data should be set as narrow as possible, as the inference is largely affected by the prior interval and its length. Improper prior could even make the algorithm miss the true posterior \cite{ref:abcsysbio}.

How to specify prior can be tricky from problem to problem. For our task in the parameter inference of regeneration model, biologically motivated priors were used, which were a general reflection of biological constraints. Also, the units of the parameters helped us in identifying the possible scale of the parameters.

@@

This subsection aims to explore options in ABC SMC implementations preliminarily. From the above results, we adopted multivariate normal kernel as the kernel for our real inference on observed data; Local M-NN kernels were also planned to be tested. In the real inference, 25:75 factor were also tried to see if it could help in recovering more features and resulting in a better fit. Because the improvement by using less data was inapparent, we did not try to change the target summary data size. Prior intervals were explored explicitly and gradually shrunk for our models.

Some other hyperparameters, e.g. the number of generations, population size (i.e. the number of particles per generation) were not addressed in this subsection. In the real inference, population size was set to be greater than 2000 to avoid insufficient exploration of the previous population; in some case than high variance or uncertainties in the results were observed, population size as large as 20,000 were used. If the epsilon did not show a convergency trend, the run could be extended to more generations until a stable result was observed.

@@

Using the suggested options in the previous section, the target data (Figure \ref{fig:obs_data}) was prepared as input for the SMC inference framework. Regarding the result, the population size and number of generations were adjusted several times to obtain a general informative posterior which should be stable across repeated runs and avoid local optima as much as possible.

Two prior intervals ([0, 25] and [0, 75]) and two distribution shape (uniform and log-uniform) were tried in the parameter estimation of the first three models. All these ABC SMC runs used 2,000 particles per population and 30 generations.

In practice, we found that log-uniform-shaped prior gave a better fit of the model than uniform-shaped prior, and wider prior range [0, 75] does not give a more accurate result than [0, 25]. Then prior interval [0, 50] was tried, in case some parameters have a true posterior laying between [25, 50].

The results of prior distribution log-uniform [0, 50] is shown in Figure \ref{fig:result123}, which is considered to give a better fit than other prior options. For these models, the acceptance rates and epsilon path are presented in Figure \ref{fig:result123_2}, and the estimated parameters' posterior of model 3 is shown in Figure \ref{fig:para1}. Figure \ref{fig:result123} plots the simulated trajectories of the three models. One thousand particles (each particle is a parameter set) were sampled, and the mean (blue) and 25th to 75th percentile interval (grey) were calculated from the data simulated by these 1000 particles.

Before applying model comparison methods, some features of the three model can still be compared. From simulated curves, It can be seen that all the inter-quartile ranges are tight around the mean value curve, which indicates the approximated posterior are concentrated; the bell-shaped posterior distributions with a single peak (Figure \ref{fig:para1}) agrees with this. All epsilon values converge to a steady level, and model 3 has a lower final epsilon value, and consequently, its simulated trajectory is more close to the observed data (model 3, Figure \ref{fig:result123}).

The acceptance rates are fluctuating and have a gradually increasing trend for all the three models, as the acceptance rate becomes higher when the true posteriors are gradually approximated. Compared to model 1, model 2 gives a better fit for the decreasing trend in the second half of the observed data. This proves that using an exponentially decaying self-increase rate $\lambda_N$ instead of a constant term indeed help to improve the model.

The resultant simulated data from model 2 and 3 have similar trends, while the most obvious difference is that for model 3, the simulated data of $\Phi$ and $\alpha$ are more `flat'. We expected model 3 to be the best as it reached a lower threshold value, and intuitively gives a better fit. However, the fit of all these models could not be considered as a generally good representation of the biological process that we want to approximate: for tnf-$\alpha$, the simulated data were significantly biased from the observed data. A sharp peak at 4 hpl was observed in the experimental data, but none of the three models represents similar features.

@@

% [model comparison among model 1, 2, 3]
